---
title: "Day 6 Lecture"
output: html_document
---

# Today's agenda

  - catching up
  - questions
  - quick re-cap
  - sketch of "data munging" procedure
  - stage 1 munging: import/fix
  - PRACTICE
  - stage 2 munging: iterative exploration/transformation
  - introducing `tidyr` and "tidy data" concepts
  - PRACTICE
  - more plotting (colors, facets)
  - summarizing/aggregating with `dplyr` package
  - quick intro to writing functions
  - PRACTICE

# Catching up

  - UGH, SORRY
  - "make-up" possibilities
  - sketch of rest of course:
    - today: reshaping, more plotting, aggregating
    - tomorrow: polishing plots, if/else structures, loops
    - Thursday: writing functions, vectorizing
    - Friday: more practice, wish list stuff
    - Monday/Tuesday(?): review, wish list
    
# Questions?

```{r}

```

# A proposed "data munging" outline/procedure

Stage 1: initial loading/fixing
  1. Read/import data
    - `read.table()` family from base R for simple text
    - `read_*()` family from `readr` package also good, can be faster
    - functions from `haven` for STATA/SPSS
    - functions from `readxl` for (fancy) Excel files
  2. Quick peek to check
    - head()
    - summary()
  3. Fix column names (make easier to work with)
    - colnames(yourdata)
    - colnames(yourdata) <- c("vector", "of", "new", "names")
    - colnames(yourdata)[c(1:3, 7, 14:16)] <- c("can", "rename", "just", "some", "selected", "columns", "too")
  4. Find/check NAs or other special values
    - summary()
    - is.na()
  5. Fix data types of columns
    - class() 
    - as.*()
    - ifelse() to change some values
  6. Fix level labels/data coding problems
    - levels(data$factor)
    - levels(data$factor) <- c("can", "assign", "new", "labels", "to", "factors")
    - ifelse()
  7. Fix level order (if needed/obvious)
    - if you want to dictate the order of a (small-ish) number of levels: factor()
    - if you just want to pick one level to be level 1 (ie "reference level"): relevel()
    - if you want to use another column in your data to set the order of levels: reorder()

# PRACTICE #1 -- #7

# Munging Stage 2: iterative exploration, reformatting, transformation

  1. Plot something
    - ggplot()
  2. Subset
    - [ , ]
    - filter(), select() from `dplyr`
  3. Combine with other data
    - rbind(), merge() from base R
    - bind_rows(), *_join() family from `dplyr`
  4. Transform
    - assign values to columns
    - mutate() from `dplyr`
  5. Reshape
    - gather(), spread() from `tidyr`
    - fancier stuff with `reshape2` package
  6. Group/summarize
    - group_by(), summarize() from `dplyr`

# PRACTICE/REVIEW #8 -- #13

# Tidy data

Some principles/practices

  1. Get data into columns (out of column names, "complex" column values)
  2. Get a 1:1 match between a (conceptual) type of data/measure and a column
    - i.e., "long" data
  3. Use/create columns to represent groupings you care about

# Introducing `tidyr`

Plot along the way!

```{r}
my.iris <- iris

# separate()

# gather()

# spread()

```

# PRACTICE: #14 -- #17

# Grouping and aggregation with `dplyr`

```{r}
# summarize()

# group_by()


```

# Quick intro to writing functions

FUNCTIONNAME <- function(arg1, arg2, ...) {
  YOUR CODE HERE
}

```{r}

```

# PRACTICE #18 -- #20