---
title: "Day 7 lecture"
output: html_document
---

# Today's agenda

  - questions/review
  - PRACTICE: reshaping/plotting
  - boxplots
  - summarizing/aggregating with `dplyr` package
  - quick intro to writing functions
  - PRACTICE
  - tweaking plots
  - using `scale_...()` functions in `ggplot2`
    - labels
    - legends
    - scales
  - adding more layers
  - annotation on plots
  - if()
  - for() loops


# PRACTICE: #1 -- #3

# Boxplots

```{r}
ff <- french_fries

ggplot(ff, aes(treatment, potato)) + geom_boxplot()

ff.long <- gather(ff, flavor, rating, potato:painty)
head(ff.long)
ggplot(ff.long, aes(treatment, rating)) + geom_boxplot() + facet_wrap(~ flavor)

ggplot(ff.long, aes(treatment, rating)) + geom_violin() + facet_wrap(~ flavor)

ggplot(ff.long, aes(treatment, rating)) + geom_boxplot() + facet_grid(rep ~ flavor)

```

# Grouping and aggregation with `dplyr`

```{r}
# summarize()
head(ff.long)
summarize(ff.long, mean.rating = mean(rating, na.rm = TRUE))
summarize(ff.long, mean.rating = mean(rating, na.rm = TRUE),
          sd.rating = sd(rating, na.rm = TRUE),
          N = length(rating))

mean(ff.long$rating, na.rm = TRUE)

# group_by()
summarize(group_by(ff.long, treatment), mean.rating = mean(rating, na.rm = TRUE))

summarize(group_by(ff.long, treatment), mean.rating = mean(rating, na.rm = TRUE))

my.results <- summarize(group_by(ff.long, treatment, flavor), mean.rating = mean(rating, na.rm = TRUE))

my.results.df <- as.data.frame(my.results)

my.results[, 1]
class(my.results[, 1])

my.results.df[, 1]
class(my.results.df[, 1])
class(my.results.df[, 1:2])

```

# Piping

```{r}
library(magrittr)

summarize(group_by(ff.long, treatment), mean.rating = mean(rating, na.rm = TRUE))

group_by(ff.long, treatment, flavor) %>% 
  summarize(mean.rating = mean(rating, na.rm = TRUE)) %>% 
  spread(flavor, mean.rating) %>% 
  as.data.frame()

ff.long %>% group_by(treatment, flavor) %>% 
  summarize(mean.rating = mean(rating, na.rm = TRUE)) %>% 
  spread(flavor, mean.rating) %>% 
  as.data.frame()

```

# Quick intro to writing functions

FUNCTIONNAME <- function(arg1, arg2, ...) {
  YOUR CODE HERE
}

```{r}
twosds <- function(vector.of.numbers) {
  output <- sd(vector.of.numbers) * 2
  return(output) # return() not strictly necessary)
  # output # this also works, instead of return(output)
}

# twosds <- function(x) { sd(x) * 2 }

twosds(rnorm(1e5, sd = 3))
twosds(rnorm(1e5, sd = 2.3))
twosds(vector.of.numbers = rnorm(1e5, sd = 76))

```

# Aggregation & more plotting

# PRACTICE #4 -- #6

# Tweaking plots

##  Audience for plots
  - For data exploration & understanding: yourself
  - For explanation & story-telling: others

## Colors

  - http://colorbrewer2.org/

```{r}
colors()

myris <- iris

myris$plant <- paste("plant", 1:nrow(myris), sep = "")

myris <- gather(myris, myvariable, cm, 1:4)
myris <- separate(myris, myvariable, c("part", "dimension"), sep = "\\.")
myris$part <- as.factor(myris$part)
myris$dimension <- as.factor(myris$dimension)

myris.hist <- ggplot(myris, aes(cm)) + geom_histogram(aes(fill = Species)) 
myris.hist
myris.hist + scale_fill_manual(values = c("red", "blue", "green"))
myris.hist + scale_fill_manual(values = c("thistle3", "lemonchiffon1", "linen"))

# RGB
umd.red <- rgb(224, 58, 62, maxColorValue = 255)
umd.gold <- rgb(255, 225, 79, maxColorValue = 255)
umd.black <- rgb(0, 0, 0, maxColorValue = 255)

maryland.colors <- c(umd.red, umd.gold, umd.black)

myris.hist + scale_fill_manual(values = c(umd.red, umd.gold, umd.black))
myris.hist + scale_fill_manual(values = maryland.colors)

my.colors <- c("pink", "seashell", umd.gold)
myris.hist + scale_fill_manual(values = my.colors)

myris.hist + scale_fill_brewer(palette = "Set1")

```

## Labels
  - multiple ways to change labels (of axes, etc.)
    - change the data
    - use "convenience" functions
    - modify the scale

```{r}

ggplot(myris, aes(cm)) + geom_histogram()

# method #1: change your data
myris$centimeters <- myris$cm
# could also use colnames

ggplot(myris, aes(centimeters)) + geom_histogram()

# method #2: "shortcut" convenience function
ggplot(myris, aes(cm)) + geom_histogram() + xlab("centimeters")

myris.wide <- select(myris, -centimeters) %>% spread(dimension, cm)

ggplot(myris.wide, aes(x = Length, y = Width)) + geom_point() + xlab("length of the part") + ylab("width of the part")

# scale_[aesthetic name]_[scale type]
ggplot(myris.wide, aes(x = Length, y = Width)) + geom_point() + 
  scale_x_continuous(name = "length of a piece of the flower")

```

## Other scale properties
  - labels
  - breaks
  - scaling (e.g., log)

```{r}
myris.scatter <- ggplot(myris.wide, aes(x = Length, Width)) + geom_point()
myris.scatter

myris.scatter + scale_x_continuous(labels = c("two", "four", "six", "seven", "eight"), breaks = c(2, 4, 6, 7, 8))
myris.scatter + scale_x_continuous(breaks = 1:8, labels = c("two", "four", "six", "eight", "red", "green", "up", "down"), name = "weird length")

myris.scatter + scale_x_continuous(breaks = seq(1, 8, .3))
myris.scatter + scale_y_continuous(breaks = seq(0, 5, .25))

myris.scatter + scale_x_log10(breaks = seq(0, 8, 2))

```

## Legends
  - position: using theme
  - name: using scale
  - levels: changing data

```{r}
myris.scatter2 <- ggplot(myris.wide, aes(Length, Width)) + geom_point(aes(color = Species), size = 4)
set1.colors <- scale_color_brewer(palette = "Set1")
myris.scatter2 
myris.scatter2 + set1.colors

myris.scatter3 <- ggplot(myris.wide, aes(Length, Width)) + geom_point(aes(color = Species, shape = Species), size = 4)
myris.scatter3

ggplot(myris.wide, aes(Length, Width)) + geom_point(aes(color = Species, shape = part), size = 4)

myris.scatter2 + set1.colors + theme(legend.position = "top")
myris.scatter2 + set1.colors + theme(legend.position = "none")


myris.scatter2 + scale_color_brewer(labels = c("a", "b", "c"), palette = "Set1")

myris.wide2 <- myris.wide
levels(myris.wide2$Species)
levels(myris.wide2$Species) <- c("a", "b", "c")

ggplot(myris.wide2, aes(x = Length, Width)) + geom_point(aes(color = Species), size = 4) + set1.colors

ggplot(myris.wide, aes(Length, Width)) + geom_point(color = "red", size = 4)

# changing font
myris.scatter2 + scale_color_discrete(name = "Iris species") + theme(text = element_text(family = "mono", size = 20))

windowsFonts()
#quartzFonts()

windowsFonts(awesome = windowsFont("TT Comic Sans MS"),
             blackadder = windowsFont("TT Blackadder ITC Regular"),
             brush = windowsFont("Brush Script MT Italic"))
windowsFonts()

myris.scatter2 + scale_color_brewer(palette = "Set1", name = "Iris species") + theme(plot.title = element_text(family = "brush", size = 20)) + ggtitle("great title")

```

- more tips on fonts using the `extrafont` package: http://blog.revolutionanalytics.com/2012/09/how-to-use-your-favorite-fonts-in-r-charts.html
- even more hilarity: https://cran.r-project.org/web/packages/xkcd/index.html

## Themes

```{r}
# theme_minimal(), theme_bw(), etc.
myris.scatter2 + theme_minimal() + theme()
myris.scatter2 + theme_bw()
myris.scatter2 + theme_classic()

```

## axis limits

```{r}

myris.scatter + xlim(-4, 20) + ylim(-3, 20)
myris.scatter + scale_y_continuous(expand = c(0, 4))

```



## More on themes

Check out this page for LOTS of options: http://docs.ggplot2.org/current/theme.html

# Additional layers

```{r}
myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2)
myris.scatter2 + geom_vline(xintercept = 5, linetype = 3)
myris.scatter2 + geom_abline(intercept = 0, slope = 1, color = "red", linetype = 2) + 
  geom_hline(yintercept = 2.5, linetype = 2) +
  geom_vline(xintercept = 5, linetype = 3)

```

# Text geom

```{r}
ggplot(myris.wide, aes(x = Length, y = Width)) + geom_text(aes(label = plant))

```


# Annotation on plots

```{r}
myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2) + 
  annotate(geom = "text", x = 2, y = 2.7, label = "interesting threshold")

myris.scatter2 + geom_hline(yintercept = 2.5, linetype = 2) + 
  annotate(geom = "text", x = 2, y = 2.7, label = "interesting threshold") +
  annotate(geom = "rect", xmin = 5, xmax = 7, ymin = 2, ymax = 3, alpha = .5) +
  annotate(geom = "text", x = 7.5, y = 2.2, label = "<-- danger zone!", color = "red")

```

