---
title: "Day 8 Practice"
output: html_document
---

# Section 1

1. Write an if...else statement to test whether a string matches "right", and if so, print "you win!", but if the string matches "wrong", print "you lose!", and if it's neither, print "the result is inconclusive".

```{r}
mystring <- "right"
mystring <- "wrong"
mystring <- "maybe"
mystring <- NA
mystring <- c("one", "wrong")

if(mystring %in% "right") {
  print("you win!")
} else if(mystring %in% "wrong") {
  print("you lose!")
} else {
  print("the result is inconclusive")
}

myfunc <- function(mystring) {
  
  if(mystring %in% "right") {
    output <- "you win!"
  } else if(mystring %in% "wrong") {
    output <- "you lose!"
  } else {
    output <- "the result is inconclusive"
  }
  return(output)
}

myfunc("right")
myfunc("wrong")
myfunc("something else")

mystrings <- c("right", "wrong", "something else")
sapply(mystrings, FUN = myfunc)

```


2. Make a copy of the iris data. Use `ifelse()` to create a new column, where:
  - if the value of Sepal.Length is less than 5, the new column should have a value of "low"
  - if the value of Sepal.Length is greater than 6, it should have a value of "high"
  - and if it's between 5 and 6 (inclusive), it should have a value of "mid"
  - make a plot in ggplot2 to confirm that you did this right
  
```{r}
myris <- iris
myris$ordinal.length <- ifelse(myris$Sepal.Length < 5, "low",
                               ifelse(myris$Sepal.Length > 6, "high", "mid"))

myris$ordinal.length <- factor(myris$ordinal.length, levels = c("low", "mid", "high"))

ggplot(myris, aes(Sepal.Length)) + geom_histogram(aes(fill = ordinal.length))

```


# Section 2

3. Practice writing a very simple loop, that simply prints out each level of Species from the iris data.  In other words, the result of running the loop should look something like:

"Level number 1 of Species is setosa"
"Level number 2 of Species is versicolor"
"Level number 3 of Species is virginica"

```{r}
# stepwise construction of the loop

# what I want:
# print(paste("Level number", SPECIESNUMBER, "of Species is", SPECIES))
# how do I loop over SPECIES and SPECIESNUMBER?
levels(iris$Species)
levels(iris$Species)[1]

for(levelnum in 1:length(levels(iris$Species))) {
   print(levelnum)
}

for(levelnum in 1:length(levels(iris$Species))) {
   print(paste("Level number", levelnum))
}

for(levelnum in 1:length(levels(iris$Species))) {
  this.level <- levels(iris$Species)[levelnum]
  print(paste("Level number", levelnum, "of Species is", this.level))
}

# with "cat" instead of "print"
for(levelnum in 1:length(levels(iris$Species))) {
  this.level <- levels(iris$Species)[levelnum]
  cat(paste("Level number", levelnum, "of Species is", this.level, "\n"))
}

# looping through strings
for(this.level in levels(iris$Species)) {   
  levelnum <- which(levels(iris$Species) %in% this.level)
  print(paste("Level number", levelnum, "of Species is", this.level))
}


```

4. Now using the loop from #3 as a starting point, for each pass through the loop, save a PDF of a scatterplot of Sepal.Length by Sepal.Width, but the plot should be different each time through the loop in the following ways:
  - each plot should only plot the subset of data from one species
  - each plot should display the plotted species as the plot title (either use `labs()` or `ggtitle()`)
  - each plot should be saved as a separate file, with the species in the file name
  - how could you change this code to write all three plots out to the same file?

```{r}
# ggtitle() example
ggplot(iris, aes(Sepal.Width)) + geom_histogram() + ggtitle("All species")

# do something once
pdf("all species plot.pdf")
ggplot(iris, aes(Sepal.Width)) + geom_histogram() + ggtitle("All species")
dev.off()

# construct the loop
for(levelnum in 1:length(levels(iris$Species))) {
  this.level <- levels(iris$Species)[levelnum]
  # print(paste("Level number", levelnum, "of Species is", this.level))
  this.data <- iris[iris$Species %in% this.level, ]
  # this.data <- filter(iris, Species %in% this.level)  # dplyr way of subsetting also fine here
  pdf(paste(this.level, "plot.pdf"))
  print(ggplot(this.data, aes(Sepal.Width)) + geom_histogram() + ggtitle(this.level))
  dev.off()
}

# multiple page PDF
pdf("species plots.pdf")

for(levelnum in 1:length(levels(iris$Species))) {
  this.level <- levels(iris$Species)[levelnum]
  # print(paste("Level number", levelnum, "of Species is", this.level))
  this.data <- iris[iris$Species %in% this.level, ]
  # this.data <- filter(iris, Species %in% this.level)  # dplyr way of subsetting also fine here
  print(ggplot(this.data, aes(Sepal.Width)) + geom_histogram() + ggtitle(this.level))
}

dev.off()

```
