---
title: "Day 8 Lecture"
output: html_document
---

# Today's agenda
  - questions/review
  - if/else
  - for-loops
    - counters (sets)
    - "growing" results
    - debugging
    - break, next
  - simulating data
  - "roll your own" power analysis

# Questions/review?

```{r}
mydata <- data.frame(n = 1:10,
                     measure = rnorm(10),
                     myfactor = c(rep("pre", 4), rep("post", 6)),
                     stringsAsFactors = FALSE)

head(mydata)
summary(mydata)

mydata$myfactor2 <- as.factor(mydata$myfactor)
summary(mydata)
levels(mydata$myfactor2)

mydata$myfactor2 <- factor(mydata$myfactor, levels = c("pre", "post"))
summary(mydata)
levels(mydata$myfactor2)

mydata$myfactor3 <- factor(mydata$myfactor, levels = c("pre", "post", "something else"))

mydata$myfactor4 <- factor(mydata$myfactor, levels = c("pre", "pos"))

summary(mydata)

mydata2 <- data.frame(year = rep(2005:2010, 3),
                      country = rep(c("USA", "Afghanistan", "Mexico"), each = 6))
mydata2

mydata2 <- expand.grid(2008:2010, c("USA", "Afghanistan", "Mexico"))
colnames(mydata2) <- c("year", "country")
mydata2$income <- rnorm(nrow(mydata2))
mydata2

mydata2.wide <- spread(mydata2, country, income)
mydata2.wide

gather(mydata2.wide, country, income, Afghanistan:Mexico)
gather(mydata2.wide, country, income, USA:Mexico)

```


# if/else
  - form: if(LOGICAL TEST) { CODE BLOCK if TEST == TRUE  }
  -       else if(OPTIONAL ADDITIONAL TEST) { CODE BLOCK if TRUE }
  - "else" code only tested if initial "if" fails
  - NOTE: "else" MUST be on same line as "}" that closes initial "if" block

```{r}
# simple examples
x <- 6
if(x < 5) {
  print("x is small")
}

x <- 3
if(x < 5) {
  print("x is small")
} else {
  print("x must be large")
}

x <- 1:6
if(is.na(x)) {
  print("x is missing, go find an x!")
} else if(x < 5) {
  print("x is too small")
} else if(x > 5) {
  print("x is too large")
} else {
  print("x is juuuust right")
}

```

  - compare with `ifelse()`
    - arguments are: test = [LOGICAL TEST], yes = [VALUE if TEST == TRUE], no = [VALUE if TEST == FALSE]

```{r} 
x <- 1:10
ifelse(x %% 2 == 0, "even", "odd")

ifelse(x %% 2 == 0, "even", 
       ifelse(x %% 3 == 0, "divisible by three", "some other odd"))

```

# PRACTICE #1 & #2

# `for` loops
  - simple way to iterate through a set of code, via a "counter" variable
  - avoid re-writing code, when possible!

```{r}
# basic structure of a loop

# going from copy-paste to a loop

```

# Some tips for loops

```{r}
# "Poor man's progress bar"


# using get() to access objects programmatically
ggplot(iris, aes(Sepal.Width)) + geom_histogram()
ggplot(iris, aes("Sepal.Width")) + geom_histogram()
ggplot(iris, aes(get("Sepal.Width"))) + geom_histogram()

# you need explicit "print" statements inside loops
for(this.dim in colnames(iris)[1:4]) {
  
}

# also for plots!
for(this.dim in colnames(iris)[1:4]) {
  
}

```


# PRACTICE #3 & #4

# Building up to a power analysis via simulation
  - simulate fake data under some parameters
  - perform test
  - examine results of many iterations

```{r}
# quick review of d/p/q/r* family of distribution functions
?rnorm

x <- sort(rnorm(1000))
summary(x)
sd(x)
hist(x)
plot(x)
plot(dnorm(x) ~ x)
plot(pnorm(x) ~ x)
plot(qnorm(seq(0, 1, .01)) ~ seq(0, 1, .01))

2*(1-pnorm(abs(1.96)))

# simulating correlations with MASS
library(MASS) # look out for "masking" of select() from dplyr
cor.table <- matrix(c(1, .2, .2, 1), nrow = 2)
cor.table
some.data <- mvrnorm(n = 10, mu = c(1, 2), Sigma = cor.table, empirical = FALSE)
some.data
summary(some.data)
cor(some.data[, 1], some.data[, 2])
as.data.frame(some.data)


```

## Step 1: write the code for one basic "iteration" of a loop

```{r}
# simulate data


# perform test


```

## Step 2: figure out how you are going to store the results of each loop
  - "pre-allocate" an object to act as the "results container"
  - modify your one-time code to "fill in" part of that container

```{r}


```

## Step 3: put code into a loop structure and test it out with a few iterations
  - make a "progress bar" if you want

```{r}


```

## Step 4: run it "for real" with full # of iterations

```{r}


```


# Power analysis
  - set up parameters of simulation
  - run simulation loop
  - process/analyze results

```{r}


```

Going further: looping over sets of simulations!
  - loops within loops

```{r}
# simple loop within loop

# nested loops in power simulation

```

# Troubleshooting loops
  - build incrementally
  - "manually" set counter, step through code inside loop
  - if you hit a snag, use "last value" of counter to identify where you hit that snag

```{r}


```

# PRACTICE #4