---
title: "Day 8 Practice"
output: html_document
---

# Section 1

1. Write an if...else statement to test whether a string matches "right", and if so, print "you win!", but if the string matches "wrong", print "you lose!", and if it's neither, print "the result is inconclusive".

```{r}
mystring <- "right"
mystring <- "wrong"
mystring <- "maybe"

if(mystring %in% "right") {
  print("you win!")
} else if(mystring %in% "wrong") {
  print("you lose!")
} else {
  print("the result is inconclusive")
}

```


2. Make a copy of the iris data. Use `ifelse()` to create a new column, where:
  - if the value of Sepal.Length is less than 5, the new column should have a value of "low"
  - if the value of Sepal.Length is greater than 6, it should have a value of "high"
  - and if it's between 5 and 6 (inclusive), it should have a value of "mid"
  - make a plot in ggplot2 to confirm that you did this right
  
```{r}
myris <- iris
myris$ordinal.length <- ifelse(myris$Sepal.Length < 5, "low",
                               ifelse(myris$Sepal.Length > 6, "high", "mid"))
ggplot(myris, aes(Sepal.Length)) + geom_histogram(aes(fill = ordinal.length))

```


# Section 2

3. Practice writing a very simple loop, that simply prints out each level of Species from the iris data.  In other words, the result of running the loop should look something like:

"Level number 1 of Species is setosa"
"Level number 2 of Species is versicolor"
"Level number 3 of Species is virginica"

```{r}


```

4. Now using the loop from #3 as a starting point, for each pass through the loop, save a PDF of a scatterplot of Sepal.Length by Sepal.Width, but the plot should be different each time through the loop in the following ways:
  - each plot should only plot the subset of data from one species
  - each plot should display the plotted species as the plot title (either use `labs()` or `ggtitle()`)
  - each plot should be saved as a separate file, with the species in the file name
  - how could you change this code to write all three plots out to the same file?

```{r}
# ggtitle() example
ggplot(iris, aes(Sepal.Width)) + geom_histogram() + ggtitle("All species")

```

# Section 2

4. Similar to what we did with the correlation power simulations, write code (including a loop) that creates a power simulation for comparing the means of two normally-distributed samples.
  - compare the sample with both `t.test()` and `wilcox.test()`
  - put the p-values from both tests in each simulation result
  - compare the power of the t-test vs. the power of the Wilcoxon test
  
```{r}
# set up the parameters of the simulation
n.sim <- 1000
# what other parameters are needed?

# set up the "empty" data frame for the results
sim.results <- data.frame(# WHAT SHOULD YOUR COLUMNS BE? 
                          )

# start the loop
# each iteration of the loop should:
# 1. generate data
# 2. perform the analysis
# 3. store the results (here, p-values) in the results data frame


# examine the results
# create a column that indicates which of the p-values reached significance


# power = the proportion of "significant" results


```

