---
title: "Day 3 Lecture"
output: html_document
---

# Review Day 2 Practice & Homework

# Questions?

# R Markdown vs. scripts
Sweave: the original "reproducible document", mix of LaTeX and R

R Markdown: one of the "cutting edge" formats, especially well-integrated into RStudio, that allows mixing of R code with text, document formatting, links, graphics, etc.

The following is a code chunk, and when `knitr()` is run on this document, it will display both the code and the output.

```{r}
x <- rnorm(10000)
hist(x)  # this is a histogram

t.test(x)
```

You can also insert "inline" results, like the following:

The mean of `x` is `r mean(x)`.

More documentation on Markdown and R Markdown:
  Help > Markdown Quick Reference
  Help > RStudio Docs

# Environments and `save()`/`load()`
  - the basic R "workspace" is the Global Environment
    - aka `.GlobalEnv`
  - "save workspace image?" question
    - recommend: no
      - not reproducible
      - helps perpetuate "bad programming hygiene"
    - ends up in .RData file (no name in front of .RData)
  - *BEST PRACTICE*: 
    - save code in script/notebook
    - save data in (non-proprietary) simple formats (CSV, etc.)
    - helps transparency, reproducibility, portability
  - HOWEVER:
    - sometimes the result of an R function has non-trivial overhead (time)
    - *for convenience*, it can be handy to save R objects in a "native" format
    - `save()` then later `load()`

```{r}
x <- rnorm(100, mean = 0, sd = 2)
y <- rnorm(100, mean = 1, sd = 1)
ttest.results <- t.test(x, y)
ttest.results

# what if ttest.results took minutes/hours to compute?
save(ttest.results, file = "big important object.RData")
ttest.results
rm(ttest.results)
ttest.results
load("big important object.RData")
ttest.results

save.image() # saves the *entire* workspace in .RData
save.image(file = "workspace.RData") # allows you to specify a name for the file
                                     # not loaded by default
```

# Reading and writing data/files
## Basic concepts:
  - not "opening" files, but rather:
    - "reading" means "pulling data from a file into the workspace"
    - "writing" means "writing data from the workspace into a file"
  - approach:
    1. Read data from whatever sources you need (from local files, from internet, whatever)
    2. Work with data in R (workspace)
    3. Write out whatever files you want to
       - updated/cleaned data
       - table of results/coefficients
       - plots
       - whatever you want! (and can program)

## Weird file types
Package for reading Excel data:
  `readxl`

  - makes it easier to deal with Excel (esp .xslx files) without needing to export to CSV

Package for other stats packages (SPSS, SAS, Stata):
  `haven`

"Tidyverse" package for reading various kinds of data:
  `readr`

## Look at the options!
```{r eval=FALSE}
?read.table

mydata <- read.csv(filename, na.strings = c("NA", "N/A", "n/a", " ", "nuthin"))

?readLines
?scan
```

## Writing files

```{r eval=FALSE}
?write.table # .csv, etc.
```

Gotcha:
  - rownames!
  
```{r eval=FALSE}
write.table(mydata, file = "outputfile.txt", 
            sep = "\t",   # this makes the output tab-delimited
            row.names = FALSE)  # good for exporting data

write.csv(mydata, file = "outputfile.csv", row.names = FALSE) # good for writing CSVs
```

# Data types
   - numeric
   - logical
   - character
   - factor
   - NA
   - NULL
   - others (e.g, "Date")

## Numeric
  - numbers!
  - sometimes integers can differ from non-integers
  - `as.numeric()` forces it to be a number (not integers)
  - `1L` is an integer
  - have mean, medians

A few handy things with numbers:

```{r}
# floating point pitfalls

# "e" notation for really big numbers
1e4

# round(), floor(), ceiling()

# sample(), rnorm(), rbinom()

# the r/p/d families

# quantile()

```


## Logicals
  - TRUE and FALSE
  - if coerced to `as.numeric`, TRUE == 1, FALSE == 0
    - can take the mean of a logical vector
  - Shortcuts: T and F
    - *DANGER*: T and F can be overwritten
  - logical operators create logical vectors
    - `==`
    - `>`
    - `<`
    - `|`  This is the "long bar", meaning (inclusive) "or" (aka "pipe")
    - `&`  == "and"
    - `>=`  greater or equal
    - `!`  is NOT
    - `&&` and `||`  for more specific programming situations
    
```{r}
# basic forms

# "shortcut" forms (avoid!)

# examples of comparisons

# which()

# logicals for subsetting

# %in%

```

PRACTICE!  #0 -- #2 in the "day03_practice.Rmd" file

Try "knitting" the result!

## Characters
  - `as.character()` converts to characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
# cat(), print()

# special characters: \t, \n, "escaping"

# nchar()

# paste(), strsplit()

# stringr package

# grep(), grepl(), gsub()

```

## Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()

# levels()

# summary()

# coersion

# subsetting

# droplevels

# changing levels (labels)

# re-ordering

```

PRACTICE: #3, #4, #5

## Data manipulation excursion
### Creating (simple) new columns

```{r}
# single value

# using subsetting

```

### Using xtabs() to check coding/cross-tabs

```{r}

```

# NAs, NULLs

## NAs
  - missing data
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}
x <- c(1, 2, NA, 4, 68)
mean(x)
mean(x, na.rm = TRUE)
y <- c(NA, 4, 23, 2, 6)

cor(x, y)
cor(x, y, use = "complete.obs")
x.1 <- c(2, 4, 68)
y.1 <- c(4, 2, 6)
cor(x.1, y.1)

x == NA
is.na(x)
!is.na(x)

```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}
x <- 1/0
y <- -1/0
z <- sqrt(-4)
x
y
z

is.finite(x)
is.finite(y)
is.finite(z)
is.finite(NA)

is.infinite(x)
is.infinite(y)
is.infinite(z)
is.infinite(NA)

is.nan(x)
is.nan(y)
is.nan(z)
is.nan(NA)

is.na(x)
is.na(y)
is.na(z) # surprising to me
is.na(NA)

```

## Generic functions
```{r}
scottdata <- read.csv("scottdata/CognitionPaperModifiedData.csv")
summary(scottdata)
summary(scottdata$Corpus)
example.lm <- lm(HiSegType ~ LoSegType, data = scottdata)
summary(example.lm)
class(scottdata)
class(scottdata$Corpus)
class(example.lm)

```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# Combining data sets

## Combining data frame objects
### `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

### `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details
