---
title: "Day 3 Practice"
output: html_document
---

0. Make a "copy" of the `iris` data set in your workspace.

```{r}
myris <- iris
```

1. Create a new data frame that is the subset of rows where Petal Length is equal to or greater than 3.3.  

```{r}
myris.sub1 <- myris[Petal.Length >= 3.3, ] # why doesn't this work?

myris.sub1 <- myris[myris$Petal.Length >= 3.3, ]
summary(myris.sub1)
```

2. Create a new data frame that is the subset of rows equal to or less than the median value of Petal Width.

```{r}
petalwidth.median <- median(myris$Petal.Width)
petalwidth.median
summary(myris)
myris.medpw <- myris[myris$Petal.Width <= petalwidth.median, ]
summary(myris.medpw)
```

3. Use `sample()` to "scramble" the data set. HINT: think about how to create a vector of numbers that represents a new ordering of the current rows.

```{r}
rownumbers <- 1:nrow(myris)
set.seed(42)
scrambled.order <- sample(rownumbers, nrow(myris), replace = FALSE)
myris.scrambled1 <- myris[scrambled.order, ]
set.seed(42)
myris.scrambled2 <- myris[sample(nrow(myris), replace = FALSE), ]

identical(myris.scrambled1, myris.scrambled2)
```

4. Find the 80th %ile of Sepal Length. Subset the data set to just the observations greater than that value. Double-check that the length (`nrow()`) of the new data set makes sense.

```{r}
sepallength.80 <- quantile(myris$Sepal.Length, .8)
sepallength.80
myris.extreme <- myris[myris$Sepal.Length > sepallength.80, ]
nrow(myris)
nrow(myris.extreme)
nrow(myris.extreme)/nrow(myris)
```

