---
title: "Day 3 Lecture"
output: html_document
---

# Review Day 2 Practice & Homework

# Questions?

## Working directory

```{r}
getwd()
# setwd()

```

# R Markdown vs. scripts
Sweave: the original "reproducible document", mix of LaTeX and R

R Markdown: one of the "cutting edge" formats, especially well-integrated into RStudio, that allows mixing of R code with text, document formatting, links, graphics, etc.

The following is a code chunk, and when `knitr()` is run on this document, it will display both the code and the output.

```{r}
x <- rnorm(10000)
hist(x)  # this is a histogram

t.test(x)
```

You can also insert "inline" results, like the following:

The mean of `x` is `r mean(x)`.

More documentation on Markdown and R Markdown:
  Help > Markdown Quick Reference
  Help > RStudio Docs

# Environments and `save()`/`load()`
  - the basic R "workspace" is the Global Environment
    - aka `.GlobalEnv`
  - "save workspace image?" question
    - recommend: no
      - not reproducible
      - helps perpetuate "bad programming hygiene"
    - ends up in .RData file (no name in front of .RData)
  - *BEST PRACTICE*: 
    - save code in script/notebook
    - save data in (non-proprietary) simple formats (CSV, etc.)
    - helps transparency, reproducibility, portability
  - HOWEVER:
    - sometimes the result of an R function has non-trivial overhead (time)
    - *for convenience*, it can be handy to save R objects in a "native" format
    - `save()` then later `load()`

```{r}
x <- rnorm(100, mean = 0, sd = 2)
y <- rnorm(100, mean = 1, sd = 1)
# ttest.results <- t.test(x, y)
# ttest.results

# what if ttest.results took minutes/hours to compute?
# save(ttest.results, file = "big important object.RData")
# ttest.results
# rm(ttest.results)
# ttest.results
load("big important object.RData")
ttest.results

save.image() # saves the *entire* workspace in .RData
save.image(file = "workspace.RData") # allows you to specify a name for the file
                                     # not loaded by default
```

# Reading and writing data/files
## Basic concepts:
  - not "opening" files, but rather:
    - "reading" means "pulling data from a file into the workspace"
    - "writing" means "writing data from the workspace into a file"
  - approach:
    1. Read data from whatever sources you need (from local files, from internet, whatever)
    2. Work with data in R (workspace)
    3. Write out whatever files you want to
       - updated/cleaned data
       - table of results/coefficients
       - plots
       - whatever you want! (and can program)

## Weird file types
Package for reading Excel data:
  `readxl`

  - makes it easier to deal with Excel (esp .xslx files) without needing to export to CSV

Package for other stats packages (SPSS, SAS, Stata):
  `haven`

"Tidyverse" package for reading various kinds of data:
  `readr`

## Look at the options!
```{r eval=FALSE}
?read.table

mydata <- read.csv(filename, na.strings = c("NA", "N/A", "n/a", " ", "nuthin"))

?readLines
?scan
```

## Writing files

```{r eval=FALSE}
?write.table # .csv, etc.
```

Gotcha:
  - rownames!
  
```{r eval=FALSE}
write.table(mydata, file = "outputfile.txt", 
            sep = "\t",   # this makes the output tab-delimited
            row.names = FALSE)  # good for exporting data

write.csv(mydata, file = "outputfile.csv", row.names = FALSE) # good for writing CSVs
```

# Data types
   - numeric
   - logical
   - character
   - factor
   - NA
   - NULL
   - others (e.g, "Date")

## Numeric
  - numbers!
  - sometimes integers can differ from non-integers
  - `as.numeric()` forces it to be a number (not integers)
  - `1L` is an integer
  - have mean, medians

A few handy things with numbers:

```{r}
# floating point pitfalls
x <- 0.1 + 0.2
x
x == 0.3

# "e" notation for really big numbers
1e4

# round(), floor(), ceiling()
x2 <- round(x, 1)
x2 == 0.3

# sample(), rnorm(), rbinom()
sample(10)

?sample

sample(1:10, size = 4, replace = FALSE)
sample(1:10, size = 4, replace = FALSE)
sample(1:10, size = 12, replace = FALSE)

sample(1:10, size = 12, replace = TRUE)

rnorm(10)

rbinom(5, size = 10, prob = .2)

many.flips <- rbinom(1e6, size = 10, prob = .2)
mean(many.flips)

# the r/p/d/q families


# quantile()
x <- rnorm(1e5, mean = 2, sd = 5)
quantile(x)

quantile(x, seq(0, 1, .1))

```


## Logicals
  - TRUE and FALSE
  - if coerced to `as.numeric`, TRUE == 1, FALSE == 0
    - can take the mean of a logical vector
  - Shortcuts: T and F
    - *DANGER*: T and F can be overwritten
  - logical operators create logical vectors
    - `==`
    - `>`
    - `<`
    - `|`  This is the "long bar", meaning (inclusive) "or" (aka "pipe")
    - `&`  == "and"
    - `>=`  greater or equal
    - `!`  is NOT
    - `&&` and `||`  for more specific programming situations
    
```{r}
# operators vs. functions
2 + 2
`+`(2, 2)

# basic forms
1 > 2
1 < 2

class(1)
class(TRUE)
class(1 < 2)
class("1")

1 == 2

x <- "a string"
x == "a string"

x != "a string"

x != 2

x <- 1:5

x <= 3
x >= 3

TRUE & TRUE
TRUE & FALSE

TRUE | FALSE

x <- 1:6
x < 3
x > 4

(x > 2) & (x < 5) # parens not necessary here
! TRUE
x > 2
!x > 2
!(x > 2)

!(x > 4 | x < 3)

# "shortcut" forms (avoid!)
F
T

F <- TRUE
F # uh oh

# examples of comparisons

# which()
x < 5

set.seed(42)
x <- sample(1:10, 100, replace = TRUE)
x == 7
which(x == 7)

# logicals for subsetting

# %in%
all.values <- c("a", "b", "c", "d")
x <- sample(all.values, 100, TRUE)

try1 <- x == c("a", "b") # NOPE
try2 <- x %in% c("a", "b") # this is usually what you want

x1 <- c(1, 2, 7, 32)
x2 <- c(1, 5, 8, 32)

x1 == x2

x1 == c(1, 2)

identical()

```

PRACTICE!  #0 -- #2 in the "day03_practice.Rmd" file

Try "knitting" the result!

  - need to set "Knit directory" to the project directory!
  - can be down with the Knit dropdown

