---
title: "Homework #1"
author: "PSYC 798W, Scott Jackson"
date: "January 7, 2019"
output: 
  tufte::tufte_handout
---

```{r echo=FALSE}
library(tufte)
```

# Homework #1
**Due before 9:30:00 EST 2019-01-08**

  1. Finish with any installation/sign-up steps that you did not complete already
  2. (optional) Run through the installation process again
    - getting a friend set up
    - installing on another computer (e.g., desktop)
    - re-installing on your computer
  3. (required) Find a data set!
  
# Finding a data set
  - Take some time to find a good data set
  - For use throughout the rest of the course & the final project
  - Requirements:
    - multivariate (i.e., a bunch of records/rows, each with multiple variables/columns)
    - mix of data types (two or more of the following):
      a. continuous or semi-continuous numbers
      b. categorical factors or groupings
      c. string/text data
      d. dates or timestamps
    - interesting enough to keep you asking questions!
  - Changing data sets later is okay, but highly dispreferred 
  - *Take some time to find a good data set* -- don't just settle for the first thing you find!

# Some places to look
  - Your own work!
  - [Google](http://lmgtfy.com/?q=public+data+sets) 
  - R has lots of built-in data sets (also available in many packages)
  - You can see all the available data sets in R by typing the command:

```{r eval=FALSE}
data()
```

  - If one of the R data sets catches your eye, talk to me if you need some basic commands to take a peek at it.

# Submitting this Homework assignment
  - *Email me the data set* `r margin_note("If it's really large, send me a link or talk with me about it.")`
  - *Email me:*
    1. List of variables you're interested in, with column names, plain English descriptions, and the TYPE of data in each variable (number, factor, text, time)
    2. ONE sentence describing why this data is interesting to you
    3. THREE initial questions you think you could ask and (at least partially) address with this data. Try to make the questions as SIMPLE as possible!
  - I will reply with comments, and suggestions for changes if needed.
  - This is the ONE assignment you can "re-do" later for credit, if I ask you to do something different.
  - BEFORE 9:30am 1/8/2018 (I will check email timestamps)
  - Even if you're not 100% decided, send me the above info for some possibility, by 9:30am, to receive credit










