---
title: "R Programming for Behavioral Science: Day 1"
author: "PSYC 798W, Scott Jackson"
date: 2019-01-07
output: ioslides_presentation
---

# Outline

## What are we doing today?
  1. Introduce each other
  2. Introduce R and some reasons why the course is focused on *programming*
  3. Discuss the course goals and structure
  4. Discuss what programming is like
  5. Introduce the basic software toolbox
  6. Get things installed
  7. Set up git
  8. Discuss next steps

# Intro

## Introductions
You!

  - pair up with the person next to you
  - ask them:
    1. Their name
    2. Their major
    3. One interesting piece of trivia
    4. Software they've used in the past for stats/data analysis/graphing
    5. What they're hoping to get out of the course
  - take notes (seriously!)
  - report to the class (and listen to each other)
  - email me your notes (scottrj@umd.edu)

## Introductions {.smaller}
Me:

  - Associate Research Scientist at Center for Advanced Study of Language
  - PhD in Linguistics (psycholing, theoretical ling, prosody)
  - Light on formal training in stats/programming
    - "stats for linguists" course (ANOVA)
    - "programming for linguists" course (Perl)
  - Heavy on self-teaching
    - 11+ years using R
    - more advanced stats (mixed-effects models, Bayesian modeling, etc.)
    - LaTeX, Emacs, git, etc.
  - Current work heavy on data analysis & visualization
  - Applied work with real-world implications (aptitude testing)
  - Learned "the hard way", hoping to help smooth the learning curve for you

## What is R?
  - programming language
  - statistical software package
  - platform for data analysis
    - stats
    - graphics
    - interface with other languages (FORTRAN, C, etc.)
  - ecosystem
    - packages (> 13,600 on [CRAN](http://cran.r-project.org/web/packages/) and counting! Last year this time, it was ~ 12,000)
    - user community

## Why *programming*?
Programming gives you *superpowers*:

   - ability to do completely new things
   - ability to replicate your work
   - ability to re-use code and get more done

Cute motivational video (watch later sometime):

https://www.youtube.com/watch?v=nKIu9yen5nc

We will be learning to *program* in R

## What does it mean to "program in R"?
   - R commands are code
   - No *technical* distinction between "user" and "programmer"
   - Approaching R like a "free, command line alternative to SPSS" is limiting, frustrating
   - Learning to think about R like a programmer is liberating, empowering
   - Good code is re-usable, expandable

## Developing useful code
   - Start with limited-case code
     - specific analyses
     - isolated scripts
   - Develop more general-use code
     - functions
     - packages
   - Develop *new* things
     - make life easier
     - make life more interesting

## The R Programming Proficiency Scale | (made up completely)

1: Cookbook

2: Basic console $\leftarrow$ where many people stop

3: Scripts (and/or notebooks)

4: Writing functions $\leftarrow$ we will get to here

4+: Creating and extending classes

5: Writing packages

# Course goals and structure

## General course approach
   - emphasis on practical use and benefits
   - skill-building
   - iterative, cumulative
   - interactive, flexible $\rightarrow$ not set in stone!
   - end-goal: produce something useful/interesting for YOU
   
## Intended course pace
  - start at the VERY bottom
  - chance to revise your assumptions
  - move VERY quickly
    - more opportunities to practice
    - solidify understanding over the course
  - YOU control
    - ask questions
    - ask for additional stuff
    - ask for review

## Class session structure
Typical class split into four parts:

1. Review: go over previous homework, answer outstanding questions
2. Lecture: me talking, you interrupting (often, I hope!) to ask questions
3. Practice: you trying something based on the lecture
3. Follow-up (and expand): discuss what broke during the Practice, go deeper (if time allows)

## Assessment structure
  - Three major components:
    1. In-class Practices
    2. Homework
    3. Final project
  - Overall grading scheme:
    - A = Completed all components
    - B = Completed two components
    - C = Completed one component
    - D = Did some work but did not complete any components
    - F = Left out one or more components entirely
    
## Completing Practice
  - the Practice component is "complete" if you complete 7/10 IN-CLASS Practices
  - *de facto* attendance
  - some Practice may require code submission
  - code submissions will only count if submitted during class
  - not graded for "correctness", only completion
  - should be automatic if you come to class
  - no "excused absences", thus the 7/10 leeway
  - HIGHLY extenuating circumstances can be discussed with me, and will require hospital notes, etc.
    
## Completing Homework {.smaller}
  - the Homework component is "complete" if you complete 7/10 assignments
  - code submissions due by 9:30am sharp the next class day after Homework is assigned
  - code submissions will be either by email or by pushing to a repository
    - described and specified for each assignment, not by choice/convenience
  - time stamps on email/pushed changes will determine if it's on time
  - Homework is "graded"
    - assignment considered complete only if:
      - the objective is completed
      AND
      - the results can be replicated by the instructor
    - Homework code will be commented and given feedback by instructor
  - What if you get stuck?  **CHEAT!!!**
    - Post questions online
    - Compare notes with other students
    
## Completing the Final Project {.smaller}
  - the Final Project has several requirements:
    - proposal due 11:59 PM EST, Sunday, January 13
    - proposal revision (if needed) due by time specified by instructor
    - final code due 8:00 PM EST, Monday, January 21
    - interesting data set (see requirements for first Homework)
  - Analysis results in one or more of:
    - numerical results
    - complex object (like a regression)
    - graphical results (e.g., plots)
  - Analysis reported and described with thorough documentation of each step
    - should enable someone to replicate analysis, even without code
  - Some "non-trivial" element not explicitly covered in class
    
## Completing the Final Project
  - "Non-trivial" elements could include:
    - Data: especially messy/big/complex data
    - Analysis: significant programming aspect to analysis (e.g., simulation)
    - Results: tricky visualization, novel way of reporting results
    - Code: useful new functions that would be of use and interest to other people
  - Code will be delivered via GitHub (preferred) or emailed
    - code will accomplish what is in the proposal
    - code will enable the instructor to completely replicate the results

# Programming workflow and mentality

## Basic GUI workflow

  - GUI = Graphical User Interface
  - what you're used to


   1. Open a file with a program (Excel, SPSS, Word, etc.)
   2. Execute commands by clicking on things, typing
   3. Save results/output by modifying file(s)

## Basic R workflow {.smaller}
   1. *Start R*
      - different interfaces available to run R
   2. *Read in data*
      - "read data", not "open files"
   3. *Write & run R code to interact with data*
      - Iterative process
      - Simple or complex
   4. *Get output*
      - Results in console
      - Graphics
      - Writing results to a (simple) file
      - Fully-reproducible papers/slides/blogs/whatever
      
## Programming/analysis workflow
   1. Set goals/questions
   2. Start R
   3. Assemble stuff you need: files, data, etc.
   4. Read/format/munge data
   5. Try something
   6. Figure out what went wrong
   7. Try something else
   8. Repeat steps 3 through 7 in some order...
   
**Programmer mentality**: *break down problems into small pieces*

## Programming/analysis workflow
[Move fast and break things](http://xkcd.com/1428/)

Things to strive for:

  - experimental approach
  - quick iteration cycle
  - break problems into smaller problems
  - learn from mistakes

## Programming habits to build
  - write code incrementally
    - start small
    - make small, modular changes
  - test often
  - expect bugs (expect things to break)
  - debugging = puzzle-solving
  - copy, steal, and cheat
    - copy other people's code and methods
    - steal ideas and implementations and useful code (with attribution!)
    - find useful ways to "cheat"" and create shortcuts for yourself

# The Toolbox

## Toolbox Overview
  - R is most effective when integrating other tools
  - Three essentials:
    - *Run* code: the base R program to interpret and run code
    - *Edit* code efficiently: a good editor or *interface* to R
    - *Track* your edits: effective use of version control software

## Running R
  - Programming languages need an interpreter/compiler to run human-readable code
  - In some languages, programs are built, then compiled/interpreted/executed
  - R is designed to be *interactive*
    - immediate execution of commands
    - iterative approach
    - extremely useful for data analysis
  - The R *language* is interpreted and executed by the open-source R *software*
  - Some other versions exist ([Microsoft R Open](https://mran.revolutionanalytics.com/open/), [pqR](http://www.pqr-project.org/), [Renjin](http://www.renjin.org/), etc.)
  - We will stick with "regular" R, aka GNU R

## Interfaces
   - Bare bones: the R console
   - Better: a functional script editor (e.g., `Rgui.exe`)
   - Best: a fully-integrated text editor
     - syntax highlighting
     - supports interaction with active console
     - good editing features
     - other bells & whistles
     
## Two excellent interfaces
[RStudio](http://www.rstudio.com/): the best dedicated R editor

  - Pros:
    - free & open source
    - fairly flexible (can support different workstyles)
    - supports reproducibility through integration with `knitr` and version control (`git` and `SVN`)
    - many R-specific conveniences
    - easy to use
    - RStudio team includes several R "luminaries"
  - Cons:
    - not great as a general-purpose editor

## Two excellent interfaces
[Emacs](http://www.gnu.org/software/emacs/): one of the best text editors of all time

  - Pros:
    - free & open source
    - integrates very well with R using the ESS package
    - infinitely flexible
    - works well for virtually *any* kind of text/code editing
    - many "killer" features work even better than in RStudio
    - you can do pretty much [anything](http://xkcd.com/378/) in Emacs
  - Cons:
    - learning curve
    - nerd jokes
       
## RStudio demo

## Version control
What is version control software?

- Tracks changes in code and other files
- Provides a universal "back button" for entire projects
- Helps with collaboration on complex projects
- Promotes good habits
- Can quickly save you immense amounts of pain & suffering

## Introducing `git`
- Different version control software (VCS) packages exist
- We will use the `git` VCS:
  - extremely powerful (you will never "outgrow" it)
  - relatively low overhead -- can be useful with some very simple commands 
  - integrates nicely with both RStudio and Emacs
  - cross-platform, free, and open-source
  - opens up additional world of R packages and development on GitHub

# Installation

## Installing R
  - Go to https://www.r-project.org/
  - Follow the CRAN link
  - Select a mirror (hosting site)
  - Follow the links and instructions for your OS
  - GO!

## Installing RStudio
  - Go to https://www.rstudio.com/
  - Download the free "desktop" version for your OS
  - GO!

## (side-track: Notepad++)
  - (For Windows)
  - Before installing git, install Notepad++ (free) 
    - https://notepad-plus-plus.org/
  - It's an excellent, user-friendly, all-purpose text editor
  - Latest installation of git sets you up with a default editor
  - Np++ by far the best option of those presented

## Installing git
  - Go to https://git-scm.com/
  - Click the download button
  - Follow installation instructions
    - For Windows especially, follow "recommendations", e.g. installing the Git Bash
  - Set up `git config`
    - Open a command-line shell (or "git shell", if you installed one)
    - type: `git version` to check that it installed
    - type: `git config --global user.name "Your Name"`
    - type: `git config --global user.email your@email`

## GitHub {.smaller}
  - [GitHub](https://github.com/) is a hosting site for code (and other projects)
  - Using `git` does not require GitHub (or any other hosting site) -- it can be used completely "locally"
  - GitHub adds some convenient wrappers and interfaces, especially for collaboration
  - GitHub is a popular place for many R developers to post code, packages, etc.
  - GitHub accounts are free for "public" repos
  - If you want private repos, you can pay a small fee
  - We will use GitHub to facilitate commenting on homework
    - You will post your homework to your own GitHub repo
    - I will "fork" the repo, make comments/changes, and then submit a "pull request", which you will be able to see, comment on, and incorporate into your code
  - [BitBucket](https://bitbucket.org/) is another option
  - [Here's a video](https://www.youtube.com/watch?v=w3jLJU7DT5E) if you want more (non-technical) descriptions about what GitHub is used for
  
## Signing up on GitHub
  - Go to https://github.com/
  - Follow the instructions to sign up
  - For convenience (not necessity), use the email you set in `git config`
    
  
## Testing and troubleshooting
  - Start "vanilla" R GUI
    - Blue console
    - Bare bones
  - Open RStudio
  - Type something in the console, like
  
```{r eval=FALSE}
2 + 2
```

## Testing version control {.smaller}
  - If you haven't already, `git config` for name and email
  - In RStudio, go to File > New Project...
    - Version Control
    - Git
    - https://github.com/scottrjackson/r_programming_bsos_winter2019.git

## Setting up a new version-controlled project {.smaller}
  - In RStudio:
    - File > New Project...
    - New Directory (or Existing Directory)
    - Navigate/name new folder
    - Select "git repository"
  - On GitHub:
    - Log in to your account
    - Click on the "+" in the upper right, select "New repository"
    - Give it the same name as your project folder from above
    - Select Public (or Private, if you don't mind paying)
    - Click "Create repository"
    
## Setting up a new version-controlled project (part two) {.smaller}
  - Back in RStudio:
  - Click the "Git" tab in the upper right
  - Select all the files (they should turn to green "A"s)
  - Click "Commit"
  - Write a commit message (like "first commit")
  - Click the "gear" icon in the Git tab (main RStudio interface)
  - Select "shell"
  - Follow the GitHub directions for "push an existing repository from the command line"
    - `git remote add origin https://github.com/USERNAME/REPONAME.git`
    - `git push -u origin master`
  - Now project can be easily committed/pushed/pulled using the RStudio interface

# Next steps

## Practices and Homework
  - Practices:
    - in-class
    - structured
    - everyone does the same thing
    - reviewed and discussed immediately
  - Homework:
    - extension of Practice/Lectures/Readings
    - on your OWN data
    - everyone does something slightly different
    - evaluated based on whether you completed the task for your data
    - collaboration and "cheating/stealing" encouraged!

## Homework and final project
  - Your final project will be a full-blown reproducible report of an interesting analysis on some data of your choosing.
  - It will represent a cumulative application of what we cover in the course.
  - You will extend it in some way that is (probably) not covered directly in class.
  - The homework assignments are designed to help you build towards this goal.

## Today's homework

See PDF in `day01` folder of the class repo for lots of details

  1. Finish with any installation/sign-up steps that you did not complete already
  2. (optional) Run through the installation process again
    - getting a friend set up
    - installing on another computer (e.g., desktop)
    - re-installing on your computer
  3. (required) Find a data set!
  
## Finding a data set {.smaller}
  - Take some time to find a good data set
  - For use throughout the rest of the course & the final project
  - Requirements:
    - multivariate (i.e., a bunch of records/rows, each with multiple variables/columns)
    - mix of data types (two or more of the following):
      - continuous or semi-continuous numbers
      - categorical factors or groupings
      - string/text data
      - dates or timestamps
    - interesting enough to keep you asking questions!
  - Changing data sets later is okay, but highly dispreferred 
  - (see the first bullet!)

## Some places to look
  - Your own work!
  - [Google](http://lmgtfy.com/?q=public+data+sets) 
  - R has lots of built-in data sets (also available in many packages)
  - You can see all the available data sets in R by typing the command:

```{r eval=FALSE}
data()
```

  - If one of the R data sets catches your eye, talk to me if you need some basic commands to take a peek at it.

## Send it to me {.smaller}
  - Email me the data set
  - If it's really large, send me a link
  - Email me:
    - List of variables you're interested in, with column names, plain English descriptions, and the TYPE of data in each variable (number, factor, text, time)
    - ONE sentence describing why this data is interesting to you
    - THREE initial questions you think you could ask and (at least partially) address with this data. Try to make the questions as SIMPLE as possible!
  - I will reply with comments, and suggestions for changes if needed.
  - This is the ONE assignment you can "re-do" later for credit, if I ask you to do something different.
  - BEFORE 9:30am 1/8/2018 (I will check email timestamps)
  - Even if you're not 100% decided, send me the above info for some possibility, by 9:30am, to receive credit

# FIN (questions?)
