---
title: "Day 4 Practice"
output: html_document
---

0. Make a copy of the `iris` data set.  Use `sample()` to make an additional "scrambled" version of the data set. 

```{r}
my.iris <- iris
my.iris.scrambled <- my.iris[sample(nrow(my.iris)), ]
head(my.iris)
head(my.iris.scrambled)
summary(my.iris)
summary(my.iris.scrambled)

```

1. Using `summary()`, identify a factor in the iris data

```{r}

```

2. Use subsetting (with logicals) to create a data frame of just the virginica species, then re-run `summary()` to check. Then "clean up" the Species factor so that it only has one level. Show how this works for both the original and the "scrambled" data frame.

```{r}

```

3. In the full data set (your copy), create different versions of the Species column:
   a. a numeric vector
   b. a character vector
   
```{r}

```

4. Use `xtabs()` to see how the values in these new columns line up with the original factor column.

```{r}

```

5. Create another column that is a copy of the factor column. Now, recode this new column so that the factor levels have new names. 

```{r}

```

6. Check the order of levels of Species with `levels()`. Create a new column from this factor, changing the order of the levels, using:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}

```

7. Explore coersion! Try to make a series of vectors with combinations from each of these data types:
  - numeric
  - character
  - factor
  - logical

For each combination, see what happens, by using `typeof()`, `class()` and `print()` before and after.

```{r}

```

8. Make a new data frame from scratch, with the following structure:
  - a numeric "ID" column that goes from 1 to 20, repeating twice (HINT: see `?seq`)
  - a factor "treatment" column that is "a" or "b", such that each ID matches up with both "a" and "b" once each
  - a numeric "outcome" variable that is a normally distributed variable, such that the outcomes for "b" are 1 unit higher on average 

```{r}

```

9. Now turn your ID column into a factor in the following way:
  - append "s" at the beginning of each number
  - convert into a factor
  - check the order of levels. Do you care? How might you change the order?

```{r}

```

10. Now change the "treatment" factor to replace "a" with "old practice" and "b" with "new technique". Do this carefully by first creating a new column that is a straight copy of the "treatment" column, then modifying that new column, then (using `xtabs()`) confirm that your new coding matches up with the old one.

```{r}

```

11. Now add a new column that just picks out "extreme values", aka values of "outcome" that are > the 90th percentile

```{r}

```

12. Make a copy of the `iris` data frame
  - Using `sample()` pull the data apart into two random batches
  - Put the pieces back together with `rbind()`
  - Put the pieces back together with `bind_rows()` from the `dplyr` package, creating an ID that keeps track of which piece is which
  
  
```{r}


```

13. Make a copy of the `iris` data frame
  - using [ , ] notation, pull the data frame into two sets of different columns
  - Put the pieces back together with `cbind()`
  - Put the pieces back together with `bind_cols()` from the `dplyr` package

```{r}


```

14. Run the following code to create a new small data frame with some additional information about iris species

```{r}
iris.info <- data.frame(Plant = c("setosa", "virginica", "limniris"),
                        State = c("Alaska", "Tennessee", "Louisiana"),
                        Rating = c(1, 5, 15))
```

Now merge this information with the `iris` data to create a new data frame. Using the `merge()` function, experiment with all = TRUE and all = FALSE. Also experiment with the `dplyr` join functions:
  - `inner_join()`
  - `full_join()`
  - `left_join()`
  - `right_join()`
  - `semi_join()`
  - `anti_join()`
  
```{r}


```

