---
title: "Day 4 Practice"
output: html_document
---

0. Make a copy of the `iris` data set.  Use `sample()` to make an additional "scrambled" version of the data set. 

```{r}
my.iris <- iris
my.iris.scrambled <- my.iris[sample(nrow(my.iris)), ]
head(my.iris)
head(my.iris.scrambled)
summary(my.iris)
summary(my.iris.scrambled)

```

1. Using `summary()`, identify a factor in the iris data

```{r}
summary(my.iris)
class(my.iris$Species)

```

Species is a factor. The summary gives counts at the different levels, which looks like a factor, and I can confirm that with `class()`

2. Use subsetting (with logicals) to create a data frame of just the virginica species, then re-run `summary()` to check. Then "clean up" the Species factor so that it only has one level. Show how this works for both the original and the "scrambled" data frame.

```{r}
virginicas <- my.iris[my.iris$Species %in% "virginica", ]
summary(virginicas)
virginicas <- droplevels(virginicas)
summary(virginicas)

virginicas.scrambled <- droplevels(my.iris.scrambled[my.iris.scrambled$Species %in% "virginica", ])
summary(virginicas.scrambled)
```

3. In the full data set (your copy), create different versions of the Species column:
   a. a numeric vector
   b. a character vector
   
```{r}
my.iris$SpeciesNum <- as.numeric(my.iris$Species)
my.iris$SpeciesChar <- as.character(my.iris$Species)
```

4. Use `xtabs()` to see how the values in these new columns line up with the original factor column.

```{r}
xtabs(~ Species + SpeciesNum, my.iris)
xtabs(~ Species + SpeciesChar, my.iris)
```

5. Create another column that is a copy of the factor column. Now, recode this new column so that the factor levels have new names. 

```{r}
my.iris$Species2 <- my.iris$Species
levels(my.iris$Species2) <- c("SETO", "VERS", "VIRG")
xtabs(~ Species + Species2, my.iris)
```

6. Check the order of levels of Species with `levels()`. Create a new column from this factor, changing the order of the levels, using:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}
my.iris$Species3 <- relevel(my.iris$Species, "versicolor")
levels(my.iris$Species3)
xtabs(~ Species + Species3, my.iris)

my.iris$SpeciesNewNum <- ifelse(my.iris$Species %in% "setosa", 3, 
                                ifelse(my.iris$Species %in% "versicolor", 1,
                                       ifelse(my.iris$Species %in% "virginica", 2, NA)))
my.iris$Species4 <- reorder(my.iris$Species, my.iris$SpeciesNewNum)
levels(my.iris$Species4)
xtabs(~ Species + Species4, my.iris)

my.iris$Species5 <- factor(my.iris$Species, levels = c("virginica", "setosa", "versicolor"))
levels(my.iris$Species5)
xtabs(~ Species + Species5, my.iris)

```

7. Explore coersion! Try to make a series of vectors with combinations from each of these data types:
  - numeric
  - character
  - factor
  - logical

For each combination, see what happens, by using `typeof()`, `class()` and `print()` before and after.

```{r}
c(FALSE, TRUE, 3)
class(c(FALSE, TRUE, 3))

c(FALSE, TRUE, "three")
class(c(FALSE, TRUE, "three"))

c("1", 2, 3)
class(c("1", 2, 3))

class(as.factor("one"))
c(as.factor("one"), 2, 33)
class(c(as.factor("one"), 2, 33))

c(as.factor("one"), "two", "three")
class(c(as.factor("one"), "two", "three"))

c(as.factor("one"), FALSE)
class(c(as.factor("one"), FALSE))

x <- as.factor(c("false", "true"))
x
class(x)
c(x, FALSE, TRUE)
class(c(x, FALSE, TRUE))

c("false", "true", FALSE, TRUE)
class(c("false", "true", FALSE, TRUE))

```

