---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
# quotes

x <- "a string"
x
y <- 'also a string'
y
z <- "this is a 'string' too"
z
w <- 'this thing is also a "string"'
w
# w.wrong <- "this thing is also a ""string""

# cat(), print()
print(x)
cat(x)

cat(x, file = "cat_example.txt")

sample.ttest <- t.test(rnorm(20, 0, 1), rnorm(20, 2, 1))
sample.ttest
print(sample.ttest)
printed.ttest <- print(sample.ttest)

# special characters: \t, \n, "escaping"
x <- "here's a cool result: 10"
cat(x)
y <- "here's a cool result:\t10"
cat(y)
z <- "here's a cool result:\n\n\t\t\t10"
cat(z)

cat("some\ntext")
cat("some\\ntext")
cat("some\"text")

# nchar()
nchar("hello")
length("hello")

# paste(), strsplit()
paste("hello", "there")
?paste

paste("hello", "there", sep = "")
x <- c("here", "are", "some", "strings")
x

paste("prefix", x, sep = "")

paste(x, collapse = " ")

x
y <- paste(x, "suffix", sep = "_")
y

strsplit(y, split = "_")

z <- paste(x, "suffix", sep = ".")
z

strsplit(z, split = ".")

strsplit(z, split = "\\.")

# stringr package

# grep(), grepl(), gsub()
x
grep("here", x)
grep("e", x)

grepl("here", x)

gsub("e", "E", x)
gsub("e", "", x)

longer.thing <- c("10F5657", "59J4957", "30K16238")
gsub("\\d+\\w(\\d+)", "\\1", longer.thing)

gsub("(\\d+)(\\w)(\\d+)", "\\2", longer.thing)
longer.thing2 <- c("10F5657ggjgjjg", "59J4957sfs", "30K16238eee")
gsub("(\\d+)(\\w)(\\d+)", "\\2", longer.thing2)

gsub("(\\d+)(\\w)(\\d+)", "\\1 \\3", longer.thing)
?regex



```

# Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()
x <- c(rep("a", 20), rep("b", 20), rep("c", 20))
x
summary(x)
x.factor <- factor(x)
summary(x.factor)

# levels()
levels(x.factor)

# summary()

# subsetting
sleep
mysleep <- sleep
summary(mysleep)
class(mysleep$group)

levels(mysleep$group)

mysleep.group1 <- mysleep[mysleep$group == "1", ]
summary(mysleep.group1)

# droplevels
mysleep.group1 <- droplevels(mysleep.group1)
summary(mysleep.group1)

# changing levels (labels)
mysleep <- sleep
levels(mysleep$group)

levels(mysleep$group) <- c("pre", "post")
levels(mysleep$group)
summary(mysleep)
head(mysleep)

mysleep.bad <- sleep
levels(mysleep.bad$group) <- c("2", "1")
summary(mysleep.bad)

mysleep.badgroup1 <- mysleep.bad[mysleep$group == "1", ]
summary(mysleep.badgroup1)

mysleep.badgroup1.wtf <- mysleep.bad[mysleep$group == 1, ]
summary(mysleep.badgroup1.wtf) # Scott was surprised this worked!

summary(mysleep.badgroup1$group)
as.numeric(mysleep.badgroup1$group)
as.numeric(as.character(mysleep.badgroup1$group))

colnames(mysleep)

colnames(mysleep) <- c("extra sleep", "treatment group", "subject")
head(mysleep)

# re-ordering
mysleep <- sleep
levels(mysleep$group) <- c("pre", "post")
summary(mysleep)
# levels(mysleep$group) <- c("post", "pre") # CHANGES DATA, NOT ORDER OF LEVELS)
# levels(mysleep$group)
mysleep$group2 <- mysleep$group
head(mysleep)
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre"))
summary(mysleep)
head(mysleep)
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre"))
mysleep$group2 <- factor(mysleep$group2, levels = c("post", "pre", "delayed post"))
summary(mysleep)
# mysleep$group2 <- factor(mysleep$group2, levels = c("pos", "pre", "delayed post"))
# summary(mysleep)


```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value
mysleep$experiment <- 1

mysleep$group3 <- mysleep$group
# using subsetting
summary(mysleep)

# ifelse()
mysleep$sleeplevel <- ifelse(mysleep$extra >= 1, "high", "low")
mysleep$sleeplevel <- ifelse(mysleep$extra >= 1, "high", 
                             ifelse(mysleep$extra < 1, "low", NA))
summary(mysleep)
mysleep$sleeplevel <- as.factor(ifelse(mysleep$extra >= 1, "high", "low"))
summary(mysleep)
mysleep$sleeplevel <- factor(ifelse(mysleep$extra >= 1, "high", "low"), levels = c("low", "high"))

summary(mysleep)

```

## Using xtabs() to check coding/cross-tabs

```{r}
xtabs(~ group + group2, data = mysleep)
xtabs(~ sleeplevel + extra, data = mysleep)
xtabs(~ extra + sleeplevel, data = mysleep)
xtabs(~ extra + sleeplevel + group, data = mysleep)

```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}

```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}
x <- 1/0
x
y <- -1/0
y
z <- sqrt(-4)
z

```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# PRACTICE: #7

