---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
# cat(), print()

# special characters: \t, \n, "escaping"

# nchar()

# paste(), strsplit()

# stringr package

# grep(), grepl(), gsub()

```

# Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()

# levels()

# summary()

# coersion

# subsetting

# droplevels

# changing levels (labels)

# re-ordering

```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value

# using subsetting

```

## Using xtabs() to check coding/cross-tabs

```{r}

```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}

```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}

```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)

# PRACTICE: #7

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# PRACTICE: #8 -- #11

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details



# PRACTICE: #12 -- #14


