---
title: "Day 9 Lecture"
output: html_document
---

# Today's agenda
  - More on loops
  - More on functions
  - Vectorization

# Loops

## Using a loop for a repetitive task

```{r}
scottdata <- read.csv("scottdata/CognitionPaperFinalData.csv")
head(scottdata)
colnames(scottdata)

# standardize every "entropy" column using scale()

# without a loop
# scottdata$AllEntropy <- as.numeric(scale(scottdata$AllEntropy))
# scottdata$WithinEntropy <- as.numeric(scale(scottdata$WithinEntropy))
# scottdata$SegEntropy <- as.numeric(scale(scottdata$SegEntropy))
# scottdata$BiphoneEntropy <- as.numeric(scale(scottdata$AllEntropy))

# with a loop
entropy.columns <- c("AllEntropy", "WithinEntropy", "SegEntropy", "BiphoneEntropy")
for(column.name in entropy.columns) {
  scottdata[, column.name] <- as.numeric(scale(scottdata[, column.name]))
}

summary(scottdata[, entropy.columns])
sapply(scottdata[, entropy.columns], sd)

```

## Using a loop with a list as a "results container"

```{r}
# use a loop to get several regression results, one for each species
my.iris <- iris

# "by hand", without a loop
lm.setosa <- lm(Sepal.Width ~ Sepal.Length, data = my.iris[my.iris$Species == "setosa", ])
lm.versicolor <- lm(Sepal.Width ~ Sepal.Length, data = my.iris[my.iris$Species == "versicolor", ])
lm.virginica <- lm(Sepal.Width ~ Sepal.Length, data = my.iris[my.iris$Species == "virginica", ])

# with a loop
all.species <- unique(as.character(my.iris$Species))
# all.species <- "some other value" # alternative

# set up container to "store" results
byspecies.results <- list()
length(byspecies.results) <- length(all.species)
names(byspecies.results) <- all.species

# loop over species names
for(this.species in all.species) {
  # run the analysis, plug the results into the "results container"
  byspecies.results[[this.species]] <- lm(Sepal.Width ~ Sepal.Length, data = my.iris[my.iris$Species == this.species, ])
}

length(byspecies.results)
class(byspecies.results[1])
class(byspecies.results[[1]])
summary(byspecies.results[[1]])
lapply(byspecies.results, summary)

```

## Extended example: power analysis

  - "power": likelihood of getting a target result
    - e.g. likelihood of "detecting" an effect at p < .05
  - process:
    1. implement data generation process
    2. simulate values using this process
    3. loop over many simulations, recording the "result" of each simulation
    4. summarize/analyze the simulated results
    
First: "type I" error, false positive rate

```{r}
# generating random (uncorrelated) data

# code up "one time" analysis

# create "results container"

# adapt "one time" code into a loop


# analyze results

```

Now: adapting to power analysis

```{r}
# generating correlated data with MASS package

# code up "one time" analysis

# create "results container"

# adapt "one time" code into a loop


# analyze results


```

Nesting loops: analyzing power under different parameters

```{r}


```

# Some other notes on loops

## Different kinds of "counters" possible

```{r}
for(counter in list(iris$Sepal.Length, iris$Sepal.Width, iris$Petal.Length, iris$Petal.Width)) {
#  print(counter)
  print(length(counter))
  print(mean(counter))
}

for(column.name in colnames(iris)[1:4]) {
  print(mean(iris[, column.name]))
}

for(counter in 1:1000) { # pretend that counter == 843 is bad, we want to skip it
  if(counter %in% c(843)) {
    
  } else {
    # whole code block here 
  }
}

```

## "Debugging" tips for for-loops

  - if you get an error, the counter variable is at the value when things broke
  - insert `print()` statements to check on things mid-loop
  - manually set the counter variable value, step through the loop

===========
# Functions
===========
  - making a function
  - anonymous functions
  - environments
  - making functions available
  - debugging

## Review function basics
  - basic syntax very simple!

```{r eval = FALSE}
my.function <- function(arg1, arg2, ...) {
  #FUNCTION BODY  
}

```

## Transitioning from a loop to a function

```{r}


```


# Vectorization overview
  - sometimes faster (not always)
  - sometimes easier to think about (not always)
  - split-apply-combine
  - "anonymous" functions = "one-use" functions

```{r}
x <- 1:10

x * 3 # "true" vectorization

x <- 1:1e6
system.time(y <- x * 3)

y2 <- vector(length = length(x))
system.time(for(i in x) {
  y2[i] <- x[i] * 3
})

```

# The `apply` family

### `sapply`
  - "simple" apply
  - returns a vector
  - good for 

```{r}

```


## `apply`

- good for "row-wise" operations over data frames (or column-wise)
- give it only the columns/rows you want to operate on
- the `MARGIN` argument says whether it's row-wise (=1) or column-wise (=2)
- the `FUN` argument says what to do with each row/column-wise vector

```{r}


# anonymous

```

## other members of the `apply` family
  - generally avoid, if possible!

### `lapply`
  - "list" apply
  - does something to each member of a list/vector
  - returns a list

### `tapply`
  - "table" apply
  - does something to data in "cells" by a combination of factors
  - returns a table


# More general solutions: `plyr` and `dplyr`
  - best for data frames
  
## `plyr`
  - the `plyr` package is older, slower, not under very active development
  - instantiates a "split-apply-combine" philosophy
  - handy paper here: http://www.jstatsoft.org/v40/i01
  - series of related functions
    - XYply, where X = "thing you give it" and Y = "thing it gives you back"
    - `ldply()` = you give it a list, it gives you back a data frame
    - `daply()` = you give it a data frame, it gives you back an array
    - etc.

## `dplyr`
  - the most common usage for `plyr` is `ddply()` -- give it a data frame and get a data frame back
  - the `dplyr` package takes this and runs with it
  - DON'T library both packages in the same session (unless you really need to and know what you're doing)
  - `dplyr` is implemented to be much faster
  - `dplyr` has rapidly taken over from `plyr`
  
### `dplyr` usage
  
  - "verbs": `mutate`, `filter`, `select`, `summarise` (or `summarize`), `arrange`
  - use `group_by()` to set the groups over which something (like `summarize` will apply)
  - `mutate` = "transform"
  - `filter` = "get some of the rows"
  - `select` = "get some of the columns"
  - `summarize` = boil things down to some kind of summary
  - `do` = general purpose!

### reviewing `summarize()`

```{r}
mysleep <- sleep
mysleep.means <- group_by(mysleep, group) %>% summarize(mean = mean(extra), sd = sd(extra))
mysleep.means
summarize(mysleep, mean = mean(extra), sd = sd(extra))

```

### `mutate()` and `rowwise()`

```{r}
mysleep$new <- mysleep$extra * 3
head(mysleep)
double.max(mysleep.wide$pre)

mysleep.wide <- mutate(mysleep.wide, newcol3 = double.max(c(pre, post)))
head(mysleep.wide)

mysleep.wide <- 
  rowwise(mysleep.wide) %>%
  mutate(newcol3 = double.max(c(pre, post))) # or newcol3 = max(pre, post) * 2
mysleep.wide

# dplyr with reshaping
mysleep.wide %>%
  gather(condition, extra.sleep, pre:post) %>%
  group_by(ID) %>%
  mutate(newcol4 = double.max(extra.sleep)) %>%
  spread(condition, extra.sleep)

```

### example of `do()` vs. a loop
  
```{r}
# using the iris data, write code to:
# - run a t.test, and 
# - pull out the t, df, and p-value into a (one-row) data frame
my.iris <- iris
ttest.results <- t.test(my.iris[, "Sepal.Length"], my.iris[, "Sepal.Width"])
t.dval <- ttest.results$statistic
df.val <- ttest.results$parameter
p.val <- ttest.results$p.value

results.table <- data.frame(t = t.val,
                            df = df.val,
                            p.val = p.val)
results.table

# write the code to do the same thing, once for "setosa" and once for "versicolor"
ttest.results <- 
  t.test(my.iris[my.iris$Species == "setosa", "Sepal.Length"], 
         my.iris[my.iris$Species == "setosa", "Sepal.Width"])
t.val <- ttest.results$statistic
df.val <- ttest.results$parameter
p.val <- ttest.results$p.value

results.table <- data.frame(t = t.val,
                            df = df.val,
                            p.val = p.val)
results.table

# write a loop to go through all levels of Species
# 1. copy & paste 1-time code
# 2. find the place(s) where the code needs to vary in each loop
# 3. replace those varying places with a counter
# 4. set up the counter to iterate through the values you want
# 5. you may need to set up a results "container" where each loop will
#    add something
# 6. double-check stuff

all.species.names <- levels(my.iris$Species)
results.table <- data.frame(Species = all.species.names,
                            t = NA,
                            df = NA,
                            p.val = NA)
results.table
for(species.name in all.species.names) {
  ttest.results <- 
    t.test(my.iris[my.iris$Species == species.name, "Sepal.Length"], 
           my.iris[my.iris$Species == species.name, "Sepal.Width"])
  t.val <- ttest.results$statistic
  df.val <- ttest.results$parameter
  p.val <- ttest.results$p.value

  results.table[results.table$Species == species.name, 
                c("t", "df", "p.val")] <- c(t.val, df.val, p.val)
}
results.table

# write a function that will return the one-row data frame of results
get.ttest <- function(data, col1, col2) {
  # data is a data frame
  # col1 is a string that's a column name
  # col2 is a string that's another column name
  data <- as.data.frame(data)
  ttest.results <- t.test(data[, col1], data[, col2])
  t.val <- ttest.results$statistic
  df.val <- ttest.results$parameter
  p.val <- ttest.results$p.value
  
  results.table <- data.frame(t = t.val,
                              df = df.val,
                              p.val = p.val)
  results.table
}
# run the function on the overall iris data

get.ttest(data = my.iris, col1 = "Sepal.Width", col2 = "Sepal.Length")

# use group_by() and do() to get the results separately for each level of Species

my.byspecies.results <- group_by(my.iris, Species) %>%
  do(get.ttest(., col1 = "Sepal.Width", col2 = "Sepal.Length")) 

my.byspecies.results

debug(get.ttest)
# undebug(get.ttest)


group_by(my.iris, Species) %>%
  do(get.ttest(., col1 = "Petal.Width", col2 = "Petal.Length")) 

head(minpair)

group_by(minpair, Corpus) %>%
  do(get.ttest(., col1 = "WithinMinPairs", col2 = "BetweenMinPairs"))

```

# Retrospective construction of the function

```{r eval = FALSE}
# start with the results I want
my.ttest.results <- data.frame(mean1 = this.mean1, mean2 = this.mean2, t = this.t, df = this.df, p = this.pval)

# how do I get those?
this.mean1 <- this.ttest$estimate[1]
this.mean2 <- this.ttest$estimate[2]
this.t <- this.ttest$statistic
this.df <- this.ttest$parameter
this.pval <- this.ttest$p.value

# where do I get the t-test?
this.ttest <- t.test(data[[var1]], data[[var2]])

# example: get it to run ONCE
this.ttest <- t.test(mydata$conditionA, mydata$conditionB)
this.mean1 <- this.ttest$estimate[1]
this.mean2 <- this.ttest$estimate[2]
this.t <- this.ttest$statistic
this.df <- this.ttest$parameter
this.pval <- this.ttest$p.value
my.ttest.results <- data.frame(mean1 = this.mean1, mean2 = this.mean2, t = this.t, df = this.df, p = this.pval)

my.ttest.results

# functionizing: make things abstract, pull out constants
# new.function <- function(v1, v2) {
# 
# ...

# new.function <- function(data) {
#  this.ttest <- t.test(data$conditionA, data$conditionB)
  
final.version <- function(data, vars) {
  this.ttest <- t.test(data[[vars[1]]], data[[vars[2]]])
  
```


# More examples

```{r}
myris <- iris
myris$plantID <- 1:nrow(myris)
myris <- gather(myris, myvariable, cm, 1:4)
myris <- separate(myris, myvariable, c("part", "dimension"), sep = "\\.")
myris$part <- as.factor(myris$part)
myris$dimension <- as.factor(myris$dimension)

head(myris)

# how many times now have we done this?
# why not make this a function?
clean.iris <- function(iris = iris) {
  library(tidyr)
  myris <- iris
  myris$plantID <- 1:nrow(myris)
  myris <- gather(myris, myvariable, cm, 1:4)
  myris <- separate(myris, myvariable, c("part", "dimension"), sep = "\\.")
  myris$part <- as.factor(myris$part)
  myris$dimension <- as.factor(myris$dimension)
  myris
}

myris2 <- clean.iris(iris)
identical(myris, myris2)

myris3 <- clean.iris(iris)

```

```{r}
# run a regression 
library(tidyverse)
head(starwars)

lm.results <- lm(get("mass") ~ height, data = starwars)
lm.summary <- summary(lm.results)
lm.summary$coefficients
# extract the estimate, p-value, and adjusted r-squared
results.table <- data.frame(estimate = lm.results$coefficients[2],
                            p.val = lm.summary$coefficients[2, "Pr(>|t|)"],
                            adj.r2 = lm.summary$adj.r.squared) 
results.table                            

# loop version: loop over genders
# NOTE: some genders only have one or two characters, so collapsing those into an "other" category
my.starwars <- starwars
my.starwars$gender2 <- ifelse(my.starwars$gender %in% c("male", "female"), my.starwars$gender, "other")
xtabs(~ gender + gender2, my.starwars)
all.genders <- unique(my.starwars$gender2)
results.table <- data.frame(gender = all.genders,
                            estimate = NA,
                            p.val = NA,
                            adj.r2 = NA)
for(this.gender in all.genders) {
  lm.results <- lm(mass ~ height, data = my.starwars[my.starwars$gender2 == this.gender, ])
  lm.summary <- summary(lm.results)
  lm.summary$coefficients
  # extract the estimate, p-value, and adjusted r-squared
  results.table[results.table$gender == this.gender, c("estimate", "p.val", "adj.r2")] <- 
    data.frame(estimate = lm.results$coefficients[2],
               p.val = lm.summary$coefficients[2, "Pr(>|t|)"],
               adj.r2 = lm.summary$adj.r.squared) 
}
results.table

# make a function to do it once
get.lm.results <- function(data, DV, predictor) {
  lm.results <- lm(get(DV) ~ get(predictor), data = data)
  lm.summary <- summary(lm.results)
  lm.summary$coefficients
  # extract the estimate, p-value, and adjusted r-squared
  results.table <- data.frame(estimate = lm.results$coefficients[2],
                              p.val = lm.summary$coefficients[2, "Pr(>|t|)"],
                              adj.r2 = lm.summary$adj.r.squared) 
  results.table               
}

get.lm.results(my.starwars, "mass", "height")

# dplyr do() version
all.results <- group_by(my.starwars, gender2) %>%
  do(get.lm.results(., "mass", "height"))
all.results

```


# More stuff on functions

## Environments
  - R can look "up" through environments, but not "down"
  - the body of a function is an environment, nested under the environment where it is called

```{r}

# remember that R can look "up" into environments
arg1 <- 40:45

bad.double.max <- function(x) { # not a good idea
  output <- max(arg1) * 2  
  output
}

y <- 1:10
bad.double.max(y)

rm(arg1)

bad.double.max(y)
y

double.max <- function(x) {
  max(x) * 2  # ok, because it's a value
}

bad.double.max2 <- function(x) {
  output <- max(x) * 2  # not returning a value
}

bad.double.max2(1:10)

#double.max <- function(x) { max(x) * 2 }
#double.max <- function(x) max(x) * 2 


double.max <- function(x) {
  output <- max(x) * 2  
  print(environment())
  print(objects())
  cat("\nThis is an awesome function\n")
  print(output)
  return(NULL)
}

x <- double.max(1:10)
x

y
double.max <- function(x) {
  output <- max(x) * 2
  y <- 6
  output
}

double.max(1:10)
y

evil.double.max <- function(x) {
  output <- max(x) * 2
  y <<- 6
  output
}

evil.double.max(5:70)
y

rm(y)
set.seed(42)
rnorm(3)
rnorm(5)
x <- evil.double.max(rnorm(20))
x
y

```


## Making functions available
  - in the notebook/script
  - `source()`
    - runs all code in another file

```{r}
# source("day9/anotherfunction.R")
```

  - package!
    - some package-making tutorials:
      - http://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/
      - http://r-pkgs.had.co.nz/
      - https://support.rstudio.com/hc/en-us/articles/200486488-Developing-Packages-with-RStudio
      - http://cran.r-project.org/doc/contrib/Leisch-CreatingPackages.pdf
      - and lots more

## Debugging
  - run as a non-function
  - insert "test" code
  - use `browser()`, `debug()`, and `undebug()`
  
```{r}
my.analysis <- function() {
  library(dplyr)
  if(require(mice)) {
    print("run mice code")
  } else {
    print("run some alternative code")
  }   
}

my.analysis2 <- function() {
  library(dplyr)
  if(library(mice)) {  # throws an error
    print("run mice code")
  } else {
    print("run some alternative code")
  }   
}


mysleep <- sleep

my.pvals.v1 <- function(data) {
  test.results <- t.test(extra ~ group, data = data, paired = TRUE)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output
}

my.pvals.v1(mysleep)
mysleep2 <- mysleep[-c(3, 17), ]
my.pvals.v1(mysleep2)
mysleep3 <- mysleep
colnames(mysleep3) <- c("time.sleep", "condition", "subject")
my.pvals.v1(mysleep3)
```

```{r eval=FALSE}
my.pvals.v2 <- function(data, DV, factor) {
  test.results <- t.test(DV ~ factor, data = data, paired = paired)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output
}

my.pvals.v2(mysleep3, DV = "time.sleep", factor = "condition") # this will break!

```

### "fake function" debugging

```{r}
my.pvals.v2(mysleep3, DV = "time.sleep", factor = "condition")

my.pvals.v2 <- function(data, DV, factor) {
  
  data <- mysleep3
  DV <- "time.sleep"
  factor <- "condition"
  
  test.results <- t.test(DV ~ factor, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output
  
  
  
}



```

### insert "print" statements

```{r}
my.pvals.v2(mysleep3, DV = "time.sleep", factor = "condition")

my.pvals.v2 <- function(data, DV, factor) {
  print(DV)
  test.results <- t.test(DV ~ factor, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output
  
  
  
}


```

### with debugging utilities
  - `browser()`
  - `debug()`, `undebug()`

```{r}
my.pvals.v2(mysleep3, DV = "time.sleep", factor = "condition")

my.pvals.v2 <- function(data, DV, factor) {
  #browser()
  test.results <- t.test(DV ~ factor, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output  
}

debug(my.pvals.v2)
undebug(my.pvals.v2)

my.pvals.v3 <- function(data, my.formula) {
  #browser()
  test.results <- t.test(my.formula, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output  
}

my.pvals.v3(mysleep3, time.sleep ~ condition)

my.pvals.v4 <- function(data, DV, factor) {
  formula.string <- paste(DV, "~", factor)
  my.formula <- as.formula(formula.string)
  test.results <- t.test(my.formula, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output  
}

my.pvals.v4(mysleep3, DV = "time.sleep", factor = "condition")


```

### dealing with user error

``` {r}
my.pvals.v4 <- function(data, DV, factor) {
  if(!is.data.frame(data)) {
    stop("Please supply a data frame, dear user") # can use warning() as well, does not throw error
  }
  formula.string <- paste(DV, "~", factor)
  my.formula <- as.formula(formula.string)
  test.results <- t.test(my.formula, data = data)
  pval <- test.results$p.value
  if (pval < .05) {
    output <- "p < .05, significant, thank god"
  } else {
    output <- paste("dammit, p is =", round(pval, 3))
  }
  output  
}

my.pvals.v4(as.matrix(mysleep3), DV = "time.sleep", factor = "condition")


```
