---
title: "Day 5 Lecture"
output: html_document
---

# Outline
  - Review Practice & Homework 4
  - Packages and namespace conflicts
  - Review NA & NULL
  - Lists
  - More on data frames
  - Simple transformations, "base" and "dplyr"
  - Simple data combinations: binding and merging
  - Graphics intro
  - REMINDER: project proposals due Sunday!

# Review

# Quick aside: package namespace conflicts

```{r}
library(MASS)
library(dplyr)

# select() from MASS is "masked"
# need to use MASS::select()

```

# More on NA, NULL

```{r}

```


# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}
# list()

# unlist()

# names() and str()

# pulling apart t-test object

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

# stringsAsFactors argument

```

# Intro to `dplyr`

```{r}

```

## subsetting with `dplyr`

```{r}

```

## "tibbles" and `as.data.frame()`

```{r}

```

# Simple transformations    

## Creating new columns

```{r}

```

## Simple column transformations

```{r}


```

## `dplyr` style: `mutate()`

```{r}

```

# PRACTICE

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` from `base` R, and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details

# PRACTICE

# Intro to graphics

http://accidental-art.tumblr.com/

## The importance of graphing

```{r}
?anscombe
example("anscombe")
```

## Two systems

  - "base" graphics
  - "grid" graphics
  - See *R Graphics* by Paul Murrell for a very thorough "deep-dive" into these implementations:  https://www.stat.auckland.ac.nz/~paul/RG2e/

### Base graphics
  - more "primitive"
  - can also be much faster than grid
  - lots of isolated plot functions
  - lots of weird little parameters to worry about
  - some handy quick-plot functions from base graphics:
    - `plot()` : a generic function that will often produce something interesting, can do scatterplots with a formula `y ~ x`
    - `hist()` : histogram of a vector
    - `boxplot()` : boxplot, using a formula `variable ~ grouping.factor`

### Grid graphics
  - better for more "high-level" plots
  - much more complex underlying system (IMHO)
  - more powerful and extendable
  - the basis of extremely popular packages such as `lattice` and `ggplot2`

## Graphics in other packages
  - many packages will have some kind of graphics functions in them, for visualizing specific kinds of things
  - these will often be in base graphics, "pre-canned", etc.
  - the two best more general-purpose graphics packages are `lattice` and `ggplot2`
  - `ggplot2` is another Hadley Wickham project, and has the most documentation, examples, and is currently most widely used

# Using `ggplot2`

## A "Grammar of Graphics"
  - based on work by Leland Wilkinson, a "theory" of graphic display of data
  - Hadley implemented this framework as his dissertation
  - `ggplot2` is the overhauled version of the original implementation, and it is still under active development
  - the idea is to implement a general system for mapping data variables to *aesthetics* (Wilkinson's term), i.e., visual dimensions (x, y, color, etc.)
  - allows for high-level control of the structure of a plot, to allow for flexible, powerful visualizations
  - the place for documentation: https://ggplot2.tidyverse.org/reference/
  - another good site: http://www.cookbook-r.com/Graphs/
  
## Basic structure and concepts

  - unlike base graphics, the *value* of `ggplot2` objects is a complex object, and `print()`-ing them renders the graph.
  - the value of the object is built up by adding *layers*
  - start with a call to `ggplot()` that specifies the data and some mappings of variables to aesthetics
  - add "geoms" to actually render something
  - add other things for fine control over graphical parameters, etc.

## Exploring data with basic geoms and aesthetics

```{r}
library(ggplot2)
?starwars

# geom_histogram()

# geom_density()

# geom_point()

# geom_text()

# geom_boxplot()

```

## More on colors
  - http://colorbrewer2.org/

```{r}
colors()
?RGB

# scale_color_brewer()

```

### Saving plots

```{r}
?Devices

```

TIP: look into other PDF readers (not Adobe) if you want to leave the file "open" (i.e., displayed) on your computer while you are editing/changing the file


