---
title: "Day 5 Lecture"
output: html_document
---

# Outline
  - Review Practice & Homework 4
  - Packages and namespace conflicts
  - Review NA & NULL
  - Lists
  - More on data frames
  - Simple transformations, "base" and "dplyr"
  - Simple data combinations: binding and merging
  - Graphics intro
  - REMINDER: project proposals due Sunday!

# Review

Couple of new handy functions

```{r}
# seq()
seq(from = 1, to = 10, by = 1)
1:10
seq(from = 20, to = 2, by = -2)
seq(from = 0, to = 1, by = 0.05)

# rep()
rep(1, 10)
rep(1:3, 5)
rep(1:3, each = 5)
rep(1:3, 5, each = 2)

```


# Quick aside: package namespace conflicts

```{r}
library(MASS)
library(dplyr)

# select() from MASS is "masked"
# need to use MASS::select()

?MASS::select
?dplyr::select # not necessary

c <- 10
c
c(1, 2, 3) # R finds the function c

c <- function(...) { print("gotcha!") }

c(1, 2, 3)

rm("c")

c(1, 2, 3)
```

# More on NA, NULL

```{r}

x <- c(1, 2, 56, NA, 43)
x
mean(x)

mysleep <- sleep
mysleep$extra[sample(1:nrow(mysleep), 5, replace = FALSE)] <- NA
summary(mysleep)

mysleep[mysleep$extra == NA, ] # does something, but not what you might expect
mysleep[is.na(mysleep$extra), ]
mysleep[!is.na(mysleep$extra), ]

?complete.cases
mean(mysleep$extra)
mean(mysleep$extra, na.rm = TRUE)
mean(!is.na(mysleep$extra), na.rm = TRUE) # getting proportion of missingness
?cor # read up on "use" argument

# read functions can help label NAs
# read.csv(..., na.strings = c("n/a", "", " ", "NA", "N/A"))

```


# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}
# list()
x <- c(1, "two", FALSE)
x
y <- list(1, "two", FALSE)
y
x[1]

y[1]
class(y[1])
y[[1]]
class(y[[1]])

# unlist()
x <- "type_instance"
strsplit(x, split = "_")
unlist(strsplit(x, split = "_"))

# names() and str()
x  <- list(1, "two", FALSE)
names(x)
names(x) <- c("first", "second", "third")
names(x)
x
x[[2]]
x[["second"]]
x["second"]
class(x)
x$second

# pulling apart t-test object
mytest <- t.test(rnorm(20, 1), rnorm(20, 2))
mytest
class(mytest)
names(mytest)
str(mytest)

mytest$p.value
mytest[["p.value"]]

# S3 vs S4 classes

# data frame are also just fancy lists
mysleep <- sleep
nrow(mysleep)
length(mysleep)
mysleep$extra

mysleep[["extra"]]
mysleep[ ,"extra"]
mysleep["extra"]

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()
mydf <- data.frame(var1 = rnorm(20, 0, 1),
                   var2 = rnorm(20, 0, 2),
                   missing = NA,
                   condition = rep(c("A", "B"), each = 2))

mydf
summary(mydf)
# stringsAsFactors argument
mydf.nofactors <- data.frame(var1 = rnorm(20, 0, 1),
                             var2 = rnorm(20, 0, 2),
                             missing = NA,
                             condition = rep(c("A", "B"), each = 2),
                             stringsAsFactors = FALSE)
summary(mydf.nofactors)


```

# Intro to `dplyr`

```{r}
library(dplyr)
```

## subsetting with `dplyr`

```{r}
mysleep <- sleep
mysleep.group2 <- mysleep[mysleep$ID %in% c("1", "4", "7"), ]
mysleep.group1 <- mysleep[mysleep$group %in% "1", ]
mysleep.group1
mysleep.group1.dplyr <- filter(mysleep, group %in% "1")
mysleep.group1.dplyr
class(mysleep.group1)
str(mysleep.group1.dplyr)

```

## "tibbles" and `as.data.frame()`

```{r}
# as.data.frame()

```

# Simple transformations    

## Creating new columns

```{r}
mysleep <- sleep
mysleep$extra2 <- mysleep$extra
head(mysleep)

# mutate() from dplyr
mysleep <- mutate(mysleep, extra3 = extra)
head(mysleep)

```

## Simple column transformations

```{r}
mysleep <- sleep
mysleep$doubleextra <- mysleep$extra * 2
mysleep$logextra <- log(abs(mysleep$extra))
head(mysleep)

# dplyr uses mutate()
mysleep <- mutate(mysleep, logextra2 = log(abs(extra)))
head(mysleep)

```

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}
mysleep <- sleep
summary(mysleep)
newsleep <- data.frame(extra = rnorm(4),
                       group = rep(c("1", "2"), each = 2),
                       ID = rep(c("11", "12"), 2))
newsleep

allsleep1 <- rbind(mysleep, newsleep)
summary(allsleep1)
summary(allsleep1$ID)
newsleep2 <- data.frame(extra = rnorm(4),
                       group = rep(c("1", "2"), each = 2),
                       ID = rep(c("11", "12"), 2),
                       gender = rep(c("M", "F"), 2))
newsleep2

allsleep2 <- rbind(mysleep, newsleep2)

# dplyr has bind_rows()
allsleep3 <- bind_rows(mysleep, newsleep2)
summary(allsleep3)


```

## `merge()` from `base` R, and `*_join()` from `dplyr` package
```{r}
mysleep <- sleep
mysleep
sleepages <- data.frame(ID = as.character(3:12), 
                        age = sample(20:50, 10),
                        gender = sample(c("M", "F", "X", NA), 10, replace = TRUE))
sleepages

sleep.merged <- merge(mysleep, sleepages) # all = FALSE
sleep.merged

sleep.merged2 <- merge(mysleep, sleepages, all = TRUE)
sleep.merged2

# dplyr way uses *_join
sleep.inner <- inner_join(mysleep, sleepages, by = "ID")
sleep.inner
sleep.outer <- full_join(mysleep, sleepages, by = "ID")
sleep.outer
sleep.left <- left_join(mysleep, sleepages, by = "ID")
sleep.left
```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details

# PRACTICE

# Intro to graphics

http://accidental-art.tumblr.com/

## The importance of graphing

```{r}
?anscombe
example("anscombe")
```

## Two systems

  - "base" graphics
  - "grid" graphics
  - See *R Graphics* by Paul Murrell for a very thorough "deep-dive" into these implementations:  https://www.stat.auckland.ac.nz/~paul/RG2e/

### Base graphics
  - more "primitive"
  - can also be much faster than grid
  - lots of isolated plot functions
  - lots of weird little parameters to worry about
  - some handy quick-plot functions from base graphics:
    - `plot()` : a generic function that will often produce something interesting, can do scatterplots with a formula `y ~ x`
    - `hist()` : histogram of a vector
    - `boxplot()` : boxplot, using a formula `variable ~ grouping.factor`

### Grid graphics
  - better for more "high-level" plots
  - much more complex underlying system (IMHO)
  - more powerful and extendable
  - the basis of extremely popular packages such as `lattice` and `ggplot2`

## Graphics in other packages
  - many packages will have some kind of graphics functions in them, for visualizing specific kinds of things
  - these will often be in base graphics, "pre-canned", etc.
  - the two best more general-purpose graphics packages are `lattice` and `ggplot2`
  - `ggplot2` is another Hadley Wickham project, and has the most documentation, examples, and is currently most widely used

# Using `ggplot2`

## A "Grammar of Graphics"
  - based on work by Leland Wilkinson, a "theory" of graphic display of data
  - Hadley implemented this framework as his dissertation
  - `ggplot2` is the overhauled version of the original implementation, and it is still under active development
  - the idea is to implement a general system for mapping data variables to *aesthetics* (Wilkinson's term), i.e., visual dimensions (x, y, color, etc.)
  - allows for high-level control of the structure of a plot, to allow for flexible, powerful visualizations
  - the place for documentation: https://ggplot2.tidyverse.org/reference/
  - another good site: http://www.cookbook-r.com/Graphs/
  
## Basic structure and concepts

  - unlike base graphics, the *value* of `ggplot2` objects is a complex object, and `print()`-ing them renders the graph.
  - the value of the object is built up by adding *layers*
  - start with a call to `ggplot()` that specifies the data and some mappings of variables to aesthetics
  - add "geoms" to actually render something
  - add other things for fine control over graphical parameters, etc.

## Exploring data with basic geoms and aesthetics

```{r}
library(ggplot2)
?starwars
sw <- starwars
# geom_histogram()
ggplot(sw, aes(mass)) + geom_histogram()

# geom_point()
ggplot(sw, aes(mass, height)) + geom_point()

# geom_text()
ggplot(sw, aes(mass, height)) + geom_text(aes(label = name))

sw2 <- filter(sw, !grepl("Jabba", name))

ggplot(sw2, aes(mass, height)) + geom_point()

ggplot(sw2, aes(mass, height)) + geom_point(aes(color = gender))

```
